God in Coq: \\ Modal Logic in Coq \\ and Scott's Version of G\"{o}del's Proof of God's Existence \\ {\small (\emph{Rough Diamond Proof Pearl})}

### 2014-01-25 Papers/ITP/GodProof-ITP.tex
Write conclusion

\newcommand{\Parameter}{\red{Parameter}}
\newcommand{\Lemma}{\red{Lemma}}
\newcommand{\Theorem}{\red{Theorem}}
\newcommand{\Definition}{\red{Definition}}
\newcommand{\Notation}{\blue{Notation}}
\newcommand{\Prop}{\blue{Prop}}
\newcommand{\Type}{\blue{Type}}
\newcommand{\iin}{\blue{in}}
\newcommand{\as}{\blue{as}}
\newcommand{\fforall}{\blue{forall}}
\newcommand{\fun}{\blue{fun}}
\newcommand{\Proof}{\blue{Proof}}
\newcommand{\Qed}{\blue{Qed}}
\newcommand{\Hint}{\blue{Hint}}
\newcommand{\com}[1]{\brown{#1}}

@@

\begin{Verbatim}[commandchars=\\\{\}]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\Parameter r: i -> i -> \Prop. \com{(* Acessibility relation for worlds *)}
\end{Verbatim}

@@

\begin{Verbatim}[commandchars=\\\{\}]
\Definition mequal (x y: u) := \fun w: i => x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).

@@

\Definition mand (p q:o)(w: i) := (p w) /\ (q w).
\Notation "p m/\ q" := (mand p q) (at level 79, right associativity).

\Definition mor (p q:o)(w: i) := (p w) \/ (q w).
\Notation "p m\/ q" := (mor p q) (at level 79, right associativity).

@@

\begin{Verbatim}[commandchars=\\\{\}]
\Definition A {t: Type}(p: t -> o) := \fun w => \fforall x, p x w.
\Notation "'mforall'  x , p" := (A (fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mforall' x : t , p" := (A (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mforall' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.

@@

\noindent
The modal operators $\Diamond$ (\emph{possibly}) and $\Box$ (\emph{necessarily}) are defined according to their meanings in the possible world semantics.

\begin{Verbatim}[commandchars=\\\{\}]
(* Modal operator for 'necessarily' *)
Definition box (p: o) := fun w => forall w1, (r w w1) -> (p w1).

(* Modal operator for 'possibly' *)
Definition dia (p: o) := fun w => exists w1, (r w w1) /\ (p w1).
\end{Verbatim}

\noindent
Although the embedding via the possible world semantics is necessary, it should be as transparent as possible to the user. Fortunately, basic Coq tactics such \texttt{intro}, \texttt{apply} and \texttt{split} automatically unfold the modal definitions and thus can be used with modal connectives and quantifiers as they are used with the usual connectives and quantifiers. Nothing else needs to be done. For the modal operators, however, the following simple tactics have been implemented, in order to allow the user to work with the concepts of \emph{necessity} and \emph{possibility} without having to unfold the definitions of modal operators and think in terms of possible worlds. These tactics remind rules of labelled calculi for modal logics.

\begin{verbatim}
Ltac box_intro w H := intros w H.

Ltac box_elim H w1 R1 Hn := 
  let P := match type of H with
             (* forall w0:i, r ?w w0 -> ?p w0 => p *)
             ((box ?p) _) => p
           end
  in cut (P w1); [intros Hn | apply (H w1 R1)].

Ltac dia_elim H w R newH := destruct H as [w [R newH]].

Ltac dia_intro w := (exists w; split; [assumption | idtac]).
\end{verbatim}

\noindent
Validity of a modal proposition is validity of its grounding on any world.

\begin{verbatim}
(* Modal validity of lifted propositions *)
Definition V (p: o) := forall w, p w.
\end{verbatim}

\noindent
The following lemmas are convenient

\begin{verbatim}
Lemma modus_ponens_inside_dia: V (mforall p, mforall q, (dia p) m-> (box (p m-> q)) m-> (dia q)).
Proof.
intro.
intros p q.
intro H1.
intro H2.
dia_elim H1 w1 R1 H1.
exists w1.
split.
  exact R1.

  apply H2. 
    exact R1.

    exact H1.
Qed.

Lemma not_dia_box_not: V (mforall p, ((m~ (dia p)) m-> (box (m~ p)) )).
Proof.
intro.
intro p.
intro H.
intros w1 H1.
intro H2.
apply H.
exists w1; split. exact H1.
exact H2.
Qed.
\end{verbatim}

\begin{verbatim}
(* Modal accessibility axioms *)

Axiom reflexivity: forall w, r w w.

Axiom transitivity: forall w1 w2 w3, (r w1 w2) -> (r w2 w3) -> (r w1 w3).

Axiom symmetry: forall w1 w2, (r w1 w2) -> (r w2 w1).
\end{verbatim}

\begin{verbatim}
(* Modal axioms *)
(* Here we show how they can be derived from the accessibility axioms *)

Theorem K: (V (mforall p, mforall q, (box (p m-> q)) m-> ( (box p) m-> (box q) ) )).
Proof.
intros w p q.
intros H1 H2.
intros w1 R1.
apply H1.
  exact R1.

  apply H2.
  exact R1.
Qed.

Theorem T: (V (mforall p, (box p) m-> p)).
Proof.
intros w p.
intro H.
apply H.
apply reflexivity.
Qed.  
\end{verbatim}

\begin{verbatim}
(* In strong modal logics, such as S5, iterations of modal operators can be collapsed *)
Theorem dia_box_to_box: V (mforall p, (dia (box p)) m-> (box p)).
Proof.
intro.
intro p.
intro H1.
destruct H1 as [w1 [R1 H1]].
intro. intro R0.
apply H1.
apply transitivity with (w2 := w).
  apply symmetry.
  exact R1.

  exact R0.
Qed.
\end{verbatim}

@@

Moreover, automated reasoning tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II}, Satallax \cite{Satallax}, Sledgehammer \cite{Sledgehammer} and Metis \cite{Hurd03first-orderproof} have been used and a few interesting new facts have been discovered about this proof \cite{TodO:ArXiv or Journal abstract}. The formalization in Isabelle \cite{ToDo:ArchiveFormalProofs}.

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable (but debatable) axioms that they should satisfy.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}).

Dana Scott's version of G\"odel's proof \cite{ScottNotes}. This formalization aims at
being as similar as possible to Dana Scott's version of the proof

The numbering of axioms, definitions and theorems is exactly the same as in Scott's notes

The formal proofs follow the same structure of Scott's proof sketches and fill their gaps

Whenever a `cut' or `assert' uses a lemma mentioned in Scott's sketches,
this is emphasized with a comment

\begin{verbatim}
Require Import Coq.Logic.Classical.
Require Import Coq.Logic.Classical_Pred_Type.

Require Import Modal.

Ltac proof_by_contradiction H := apply NNPP; intro H.

(* Constant predicate that distinguishes positive properties *)
Parameter Positive: (u -> o) -> o.

(* Axiom A1: either a property or its negation is positive, but not both *)
Axiom axiom1a : V (mforall p, (Positive (fun x: u => m~(p x))) m-> (m~ (Positive p))).
Axiom axiom1b : V (mforall p, (m~ (Positive p)) m-> (Positive (fun x: u => m~ (p x))) ).

(* Axiom A2: a property necessarily implied by a positive property is positive *)
Axiom axiom2: V (mforall p, mforall q, Positive p m/\ (box (mforall x, (p x) m-> (q x) )) m-> Positive q).

(* Theorem T1: positive properties are possibly exemplified *)
Theorem theorem1: V (mforall p, (Positive p) m-> dia (mexists x, p x) ).
Proof.
intro.
intro p.
intro H1.
proof_by_contradiction H2.
apply not_dia_box_not in H2.
assert (H3: ((box (mforall x, m~ (p x))) w)). (* Lemma from Scott's notes *)
  box_intro w1 R1.
  intro x.
  assert (H4: ((m~ (mexists x : u, p x)) w1)).
    box_elim H2 w1 R1 G2. 
    exact G2.

    clear H2 R1 H1 w.
    intro H5.
    apply H4.
    exists x.
    exact H5.

  assert (H6: ((box (mforall x, (p x) m-> m~ (x m= x))) w)). (* Lemma from Scott's notes *)    
    box_intro w1 R1.
    intro x.
    intro H7.
    intro H8.
    box_elim H3 w1 R1 G3.
    apply G3 with (x := x).
    exact H7.

    assert (H9: ((Positive (fun x => m~ (x m= x))) w)). (* Lemma from Scott's notes *)
    apply (axiom2 w p (fun x => m~ (x m= x))).
    split.
      exact H1.

      exact H6.
    assert (H10: ((box (mforall x, (p x) m-> (x m= x))) w)). (* Lemma from Scott's notes *)
      box_intro w1 R1.
      intros x H11.     
      reflexivity.

      assert (H11 : ((Positive (fun x => (x m= x))) w)). (* Lemma from Scott's notes *)
        apply (axiom2 w p (fun x => x m= x )).
        split.
          exact H1.

@@

        clear H1 H2 H3 H6 H10 p.
        apply axiom1a in H9.
        contradiction.
Qed.

(* Definition D1: God: a God-like being possesses all positive properties *)
Definition G(x: u) := mforall p, (Positive p) m-> (p x).

(* Axiom A3: the property of being God-like is positive *)
Axiom axiom3: V (Positive G).

(* Corollary C1: possibly, God exists *)
Theorem corollary1: V (dia (mexists x, G x)). 
Proof.
intro. 
apply theorem1.
apply axiom3.
Qed.

(* Axiom A4: positive properties are necessarily positive *)
Axiom axiom4: V (mforall p, (Positive p) m-> box (Positive p)).

(* Definition D2: essence: an essence of an individual is a property possessed by it and necessarily implying any of its properties *)
Definition Essence(p: u -> o)(x: u) := (p x) m/\ mforall q, ((q x) m-> box (mforall y, (p y) m-> (q y))).
Notation "p 'ess' x" := (Essence p x) (at level 69).

(* Theorem T2: being God-like is an essence of any God-like being *)
Theorem theorem2: V (mforall x, (G x) m-> (G ess x)).
Proof.
intro.
intro g.
intro H1.
unfold Essence.
split.
  exact H1.

@@

    cut (box (Positive q) w). (* Lemma from Scott's notes *)
      apply K.
      box_intro w1 R1.
      intro H5.
      intro y.
      intro H6.
      unfold G in H6.
      apply (H6 q).
      exact H5.

      apply axiom4.
      exact H3.
Qed.

(* At this point in Scott's notes there are two notes that are not necessary for the proof, *)
(* but it would be interesting to formalize them anyway *)

(* Definition D3: necessary existence: necessary existence of an individual is the necessary exemplification of all its essences *)
Definition NE(x: u) := mforall p, (p ess x) m-> box (mexists y, (p y)).

(* Axiom A5: necessary existence is a positive property *)
Axiom axiom5: V (Positive NE).

Lemma lemma1: V ((mexists z, (G z)) m-> box (mexists x, (G x))).
Proof.
intro.
intro H1.
destruct H1 as [g H2].
cut ((G ess g) w).      (* Lemma from Scott's notes *)
  assert (H3: (NE g w)).       (* Lemma from Scott's notes *)
    unfold G in H2.
    apply (H2 NE).
    apply axiom5.

@@

  apply theorem2.
  exact H2.
Qed.

Lemma lemma2: V (dia (mexists z, (G z)) m-> box (mexists x, (G x))).
Proof.
intro.
intro H.
cut (dia (box (mexists x, G x)) w).  (* Lemma from Scott's notes *)
  apply dia_box_to_box.

  apply (modus_ponens_inside_dia w (mexists z, G z)).
    exact H.

    box_intro w1 R1.
    apply lemma1.
Qed.

(* Theorem T3: necessarily, a God exists *)
Theorem theorem3: V (box (mexists x, (G x))).
Proof.
intro.
apply lemma2.
apply corollary1.
Qed.

(* Corollary C2: There exists a god *)
Theorem corollary2: V (mexists x, (G x)).
Proof.
intro.
apply T.
apply theorem3.
Qed.
\end{verbatim}

Note: possible worlds machinery is hidden from the user.

@@

cite related work about modal logics in Coq.

no need to bother about the embedding.

This work attests the maturity of contemporary interactive and
automated deduction tools for classical higher-order logic and
demonstrates the elegance and practical relevance of the
embeddings-based approach.  Most importantly, our work opens new
perspectives for a computer-assisted theoretical philosophy.  The
critical discussion of the underlying concepts, definitions and axioms
remains a human responsibility, but the computer can assist in
building and checking rigorously correct logical arguments. In case of
logico-philosophical disputes, the computer can check the disputing
arguments and partially fulfill Leibniz' dictum: Calculemus --- Let us
calculate!
The philosophical and theological importance of the correctness of this proof is beyond the scope of this paper.

We welcome feedback on

\paragraph{Acknowledgements:} ToDo: thank Coq-Club members. Cedric Auger

>>>>

\newcommand{\Parameter}{\red{Parameter}}
\newcommand{\Ltac}{\red{Ltac}}
\newcommand{\Axiom}{\red{Axiom}}
\newcommand{\Lemma}{\red{Lemma}}
\newcommand{\Theorem}{\red{Theorem}}
\newcommand{\Definition}{\red{Definition}}
\newcommand{\Notation}{\blue{Notation}}
\newcommand{\Prop}{\blue{Prop}}
\newcommand{\Type}{\blue{Type}}
\newcommand{\llet}{\blue{let}}
\newcommand{\match}{\blue{match}}
\newcommand{\with}{\blue{with}}
\newcommand{\eend}{\blue{end}}
\newcommand{\iin}{\blue{in}}
\newcommand{\as}{\blue{as}}
\newcommand{\Require}{\blue{Require}}
\newcommand{\Import}{\blue{Import}}
\newcommand{\fforall}{\blue{forall}}
\newcommand{\fun}{\blue{fun}}
\newcommand{\Proof}{\blue{Proof}}
\newcommand{\Qed}{\blue{Qed}}
\newcommand{\Hint}{\blue{Hint}}
\newcommand{\com}[1]{\brown{#1}}
\newcommand{\bslash}{\symbol{92}}

@@

\newcommand{\verbsize}{\small}

\begin{small}
\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Parameter i: \Type. \com{(* Type for worlds *)}
\Parameter u: \Type. \com{(* Type for individuals *)}
\Definition o := i -> \Prop. \com{(* Type of modal propositions *)}
\Parameter r: i -> i -> \Prop. \com{(* Acessibility relation for worlds *)}
\end{Verbatim}
\end{small}

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition mequal (x y: u) := \fun w: i => x = y.
\Notation "x m= y" := (mequal x y) (at level 99, right associativity).

@@

\Definition mand (p q:o)(w: i) := (p w) /\bslash (q w).
\Notation "p m/\bslash q" := (mand p q) (at level 79, right associativity).

\Definition mor (p q:o)(w: i) := (p w) \bslash\slash (q w).
\Notation "p m\bslash\slash q" := (mor p q) (at level 79, right associativity).

@@

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition A {t: Type}(p: t -> o) := \fun w => \fforall x, p x w.
\Notation "'mforall'  x , p" := (A (fun x => p))
  (at level 200, x ident, right associativity) : type_scope.
\Notation "'mforall' x : t , p" := (A (\fun x:t => p))
  (at level 200, x ident, right associativity, 
    format "'[' 'mforall' '/ '  x  :  t , '/ '  p ']'")
  : type_scope.

@@

\noindent
The modal operators $\Diamond$ (\emph{possibly}) and $\Box$ (\emph{necessarily}) are defined accordingly to their meanings in the possible world semantics.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition box (p: o) := \fun w => \fforall w1, (r w w1) -> (p w1).
\Definition dia (p: o) := \fun w => exists w1, (r w w1) /\bslash (p w1).
\end{Verbatim}

\noindent
Although the embedding via the possible world semantics is necessary, it should be as transparent as possible to the user. Fortunately, basic Coq tactics such \texttt{intro}, \texttt{apply} and \texttt{split} automatically unfold the modal definitions and thus can be used with modal connectives and quantifiers as they are used with the usual connectives and quantifiers. Nothing else needs to be done. For the modal operators, however, the following simple tactics have been implemented, in order to allow the user to work with the concepts of \emph{necessity} and \emph{possibility} without having to unfold the definitions of modal operators and think in terms of possible worlds.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Ltac box_i := \llet w := fresh "w" \iin \llet R := fresh "R" 
               \iin (intro w at top; intro R at top).

\Ltac box_elim H w1 H1 := \match type of H \with 
  ((box ?p) ?w) => cut (p w1); [intros H1 | (apply (H w1); try assumption)] \eend.

\Ltac box_e H H1:= \match goal \with | [ |- (_ ?w) ] => box_elim H w H1 \eend.

\Ltac dia_e H := \llet w := fresh "w" \iin \llet R := fresh "R" 
                 \iin (destruct H \as [w [R H]]; move w at top; move R at top).

\Ltac dia_i w := (exists w; split; [assumption | idtac]).
\end{Verbatim}

\noindent
Validity of a modal proposition is validity of its grounding on any world. A non-intrusive notation and a tactic that grounds the proposition to an arbitrary world are provided.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Definition V (p: o) := \fforall w, p w.
\Notation "[ p ]" := (V p).
\Ltac mv := \match goal \with [|- (V _)] => intro \eend.
\end{Verbatim}

\noindent
The following lemmas are quite convenient. The first is a kind of modus ponens applicable to formulas under the scope of a modal operator. The second allows pushing negations inside modalities.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Lemma mp_dia: [mforall p, mforall q, (dia p) m-> (box (p m-> q)) m-> (dia q)].
\Proof. mv. intros p q H1 H2. dia_e H1. dia_i w0. box_e H2 H3. apply H3. exact H1. \Qed.

\Lemma not_dia_box_not: [mforall p, (m~ (dia p)) m-> (box (m~ p))].
\Proof. mv. intro p. intro H. box_i. intro H2. apply H. dia_i w0. exact H2. \Qed.
\end{Verbatim}

\noindent
The most well-known modal logics can be distinguished by their acceptance of the following reachability axioms. \textbf{S5} is the modal logic that accepts all three.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Axiom reflexivity: \fforall w, r w w.
\Axiom transitivity: \fforall w1 w2 w3, (r w1 w2) -> (r w2 w3) -> (r w1 w3).
\Axiom symmetry: \fforall w1 w2, (r w1 w2) -> (r w2 w1).
\end{Verbatim}

\noindent
Modal logic ``axioms'' can be derived from the reachability axioms.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem K: [ mforall p, mforall q, (box (p m-> q)) m-> (box p) m-> (box q) ].
\Proof. mv. intros p q H1 H2. box_i. box_e H1 H3. apply H3. box_e H2 H4. exact H4. \Qed.

\Theorem T: [ mforall p, (box p) m-> p ].
\Proof. mv. intro p. intro H. box_e H H1. exact H1. apply reflexivity. \Qed.
\end{Verbatim}

\noindent
In strong modal logics, such as S5, iterations of modal operators can be collapsed. This is a controversial principle used in G\"odel's and Scott's versions of the proof.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Theorem dia_box_to_box: [ mforall p, (dia (box p)) m-> (box p) ].
\Proof. mv. intro p. intro H1. dia_e H1. box_i. box_e H1 H2. exact H2. eapply transitivity.
  apply symmetry. exact R.

  exact R0.
\Qed.
\end{Verbatim}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  
G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). The formalization shown in this section aims at
being as similar as possible to Dana Scott's version of the proof \cite{ScottNotes}. The formulation and numbering of axioms, definitions and theorems is the same as in Scott's notes. Even the Coq proof scripts follow all the steps in Scott's notes. Scott's assertions are emphasized with comments. In contrast to the formalization in Isabelle \cite{ToDo:our isabelle formalization in ArchiveFormalProofs}, where automation via Metis \cite{Hurd03first-orderproof} and Sledgehammer \cite{Sledgehammer} using tools such as Nitpick \cite{Nitpick}, LEO-II \cite{LEO-II} and Satallax \cite{Satallax} has been successfully employed, the formalization in Coq used no automation. This was a deliberate choice, mainly because it allowed a qualitative evaluation of the convenience of the possible world semantic embedding approach for \emph{interactive} theorem proving. Moreover, in order to formalize precisely Scott's version and not another automatically found version, automation would have to be heavily limited anyway. Furthermore, the deliberate preference for simple tactics (mostly \emph{intro}, \emph{apply} and the modal tactics described in the previous section) results in proof scripts that could be read as natural deduction proofs. This hopefully makes the formalization more accessible to those who are not experts in Coq's tactics but are nevertheless interested in G\"odel's proof.

\begin{Verbatim}[commandchars=\\\{\},fontsize=\verbsize]
\Require \Import Coq.Logic.Classical Coq.Logic.Classical_Pred_Type Modal.

\Ltac proof_by_contradiction H := apply NNPP; intro H.

\com{(* Constant predicate that distinguishes positive properties *)}
\Parameter Positive: (u -> o) -> o.

\com{(* Axiom A1: either a property or its negation is positive, but not both *)}
\Axiom axiom1a : [ mforall p, (Positive (\fun x: u => m~(p x))) m-> (m~ (Positive p)) ].
\Axiom axiom1b : [ mforall p, (m~ (Positive p)) m-> (Positive (\fun x: u => m~ (p x))) ].

\com{(* Axiom A2: a property necessarily implied by a positive property is positive *)}
\Axiom axiom2: [ mforall p, mforall q, 
  Positive p m/\bslash (box (mforall x, (p x) m-> (q x) )) m-> Positive q ].

\com{(* Theorem T1: positive properties are possibly exemplified *)}
\Theorem theorem1: [ mforall p, (Positive p) m-> dia (mexists x, p x) ].
\Proof. mv.
intro p. intro H1. proof_by_contradiction H2. apply not_dia_box_not in H2.
assert (H3: ((box (mforall x, m~ (p x))) w)). \com{(* Scott *)}
  box_i. intro x. assert (H4: ((m~ (mexists x : u, p x)) w0)).
    box_e H2 G2. exact G2.

    clear H2 R H1 w. intro H5. apply H4. exists x. exact H5.

  assert (H6: ((box (mforall x, (p x) m-> m~ (x m= x))) w)). \com{(* Scott *)}   
    box_i. intro x. intros H7 H8. box_elim H3 w0 G3. eapply G3. exact H7.

    assert (H9: ((Positive (fun x => m~ (x m= x))) w)). \com{(* Scott *)}
      apply (axiom2 w p (fun x => m~ (x m= x))). split.
        exact H1.

        exact H6.
      assert (H10: ((box (mforall x, (p x) m-> (x m= x))) w)). \com{(* Scott *)}
        box_i. intros x H11. reflexivity.

        assert (H11 : ((Positive (fun x => (x m= x))) w)). \com{(* Scott *)}
          apply (axiom2 w p (fun x => x m= x )). split.
            exact H1.

@@

          apply axiom1a in H9. contradiction.
\Qed.

\com{(* Definition D1: God: a God-like being possesses all positive properties *)}
\Definition G(x: u) := mforall p, (Positive p) m-> (p x).

\com{(* Axiom A3: the property of being God-like is positive *)}
\Axiom axiom3: [ Positive G ].

\com{(* Corollary C1: possibly, God exists *)}
\Theorem corollary1: [ dia (mexists x, G x) ]. 
\Proof. mv. apply theorem1. apply axiom3. \Qed.

\com{(* Axiom A4: positive properties are necessarily positive *)}
\Axiom axiom4: [ mforall p, (Positive p) m-> box (Positive p) ].

\com{(* Definition D2: essence: an essence of an individual is a property possessed by it} 
\com{and necessarily implying any of its properties *)}
\Definition Essence(p: u -> o)(x: u) := (p x) m/\bslash mforall q, 
                                        ((q x) m-> box (mforall y, (p y) m-> (q y))).
\Notation "p 'ess' x" := (Essence p x) (at level 69).

\com{(* Theorem T2: being God-like is an essence of any God-like being *)}
\Theorem theorem2: [ mforall x, (G x) m-> (G ess x) ].
\Proof. mv. intro g. intro H1. unfold Essence. split.
  exact H1.

@@

    cut (box (Positive q) w). \com{(* Scott *)}
      apply K. box_i. intro H5. intro y. intro H6. unfold G in H6. apply (H6 q). exact H5.

      apply axiom4. exact H3.
\Qed.

\com{(* Definition D3: necessary existence: necessary existence of an individual }
\com{is the necessary exemplification of all its essences *)}
\Definition NE(x: u) := mforall p, (p ess x) m-> box (mexists y, (p y)).

\com{(* Axiom A5: necessary existence is a positive property *)}
\Axiom axiom5: [ Positive NE ].

\Lemma lemma1: [ (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H1. destruct H1 as [g H2]. cut ((G ess g) w). \com{(* Scott *)}
  assert (H3: (NE g w)).       \com{(* Scott *)}
    unfold G in H2. apply (H2 NE). apply axiom5.

@@

  apply theorem2. exact H2.
\Qed.

\Lemma lemma2: [ dia (mexists z, (G z)) m-> box (mexists x, (G x)) ].
\Proof. mv. intro H. cut (dia (box (mexists x, G x)) w).  \com{(* Scott *)}
  apply dia_box_to_box.

  apply (mp_dia w (mexists z, G z)).
    exact H.

    box_i. apply lemma1.
\Qed.

\com{(* Theorem T3: necessarily, a God exists *)}
\Theorem theorem3: [ box (mexists x, (G x)) ].
\Proof. mv. apply lemma2. apply corollary1. \Qed.

\com{(* Corollary C2: There exists a god *)}
\Theorem corollary2: [ mexists x, (G x) ].
\Proof. mv. apply T. apply theorem3. \Qed.
\end{Verbatim}

@@

The successful formalization of Scott's version of G\"odel's proof of 
God's existence indicates that the 
\emph{embedding} of modal logics 
into higher-order logics via the \emph{possible world semantics} is a viable approach for interactive 
theorem proving within modal logics.
The minimalistic implementation of the embedding 
(Section \ref{sec:modal}) takes special care to hide the 
underlying possible world machinery from the user. 
An inspection of the proof scripts in Section \ref{sec:proof} shows 
that this goal has been achieved. The user does not have to 
explicitly bother about worlds and their mutual reachability; 
the provided tactics for modalities do the job for him. 
And for subgoals that do not involve modalities, 
the user has all the usual interactive tactics at his/her disposal. 
%The implementation of hints to allow automation tactics to fully cope with the embedding remains for future work. Nevertheless, at this stage a tactic such as \texttt{firstorder} is able to prove all lemmas and theorems in Section \ref{sec:modal} fully automatically.

The theological implications of the verified correctness of 
this proof certainly depend on a critical discussion of the 
underlying concepts, definitions and axioms, 
which is beyond the scope of this paper. Nevertheless, the application
of theorem proving technology to a philosophical domain 
has been fruitful for both areas. New results of philosophical 
interest have been obtained \cite{ToDo:our ArXiv or Journal abstract} 
and basic infra-structure for interactive and automated reasoning 
in higher-order modal logics has been created.

\marginpar{cite John Rushby's formalization of anselm's argument?}

\paragraph{Acknowledgements:} Thanks to Cedric Auger and Laurent Th\'ery, for their answers to our newbie questions about Ltac in the Coq-Club mailing-list.
